Making Sequential Programs Deterministic

### 2019-04-06 design.tex
Fix font size
Edits in 2.3

    \caption{Type well-formedness}
\todo{What if $\beta$ has \<List> within?  Be sure everything works in that case.}

@@

The element type of a \<NonDet> collection must be \<NonDet>.
Otherwise, the type qualifier on a collection's element must be a subtype or equal to
the collection's type qualifier (\rulename{collection} rules of
\cref{type-validity}). Note that the \rulename{ordernondet} rules also
permit \<Det> collections,
because $\<Det> \sqsubseteq \<OrderNonDet>$.
\Cref{fig:determinism-collections} restates the \rulename{collection} rules
of
\cref{type-validity}.

@@

\end{Verbatim}
\end{minipage}
\hfill
\begin{minipage}{.46\textwidth}
These types are invalid:
\begin{smaller}
\begin{Verbatim}[commandchars=\\\{\}]
NonDet      \lptype{List}{OrderNonDet \lptype{Set}{...}}
NonDet      \lptype{List}{Det         Int}
OrderNonDet \lptype{List}{NonDet      Int}
Det         \lptype{List}{NonDet      Int}
Det         \lptype{List}{OrderNonDet \lptype{Set}{...}}
\end{Verbatim}
\end{smaller}
\end{minipage}
\smallskip

@@

\begin{enumerate}
\item
The individual elements retrieved from it have type \<NonDet>.  This
affects access, iteration, searching, etc.
(Rules \rulename{iteration} and \rulename{next element} in \cref{fig-ordernondet-rules}.)
\item
Size-related operations return a deterministic result.  This also affects
queries of whether an iterator has more elements.
\item
If the collection is sorted, or its elements are placed in a collection
that does sorting, the result is deterministic.
\end{enumerate}

To state the first point more formally, the 
typical type for a list access operation, such as \<get>, is
$$\forallt{\tau} \<List>\angles{\tau} \times \<Det Int> \rightarrow \tau$$
but this type is incorrect for \ourTypeSystem.  Each of the following is a
correct (but overly restrictive) type for \<get>:
\begin{eqnarray*}
\forallt{\kappa\ \beta} \<NonDet>\ \<List>\angles{\kappa\ \beta} \times \<Det Int>
& \strut\hspace{-1em} \rightarrow \<NonDet>\ \beta
& \mbox{\quad in this case $\kappa = \<NonDet>$} \\
\forallt{\kappa\ \beta} \<OrderNonDet>\ \<List>\angles{\kappa\ \beta} \times \<Det Int>
& \strut\hspace{-1em} \rightarrow \<NonDet>\ \beta
& \mbox{\quad in this case $\kappa \in \{ \<OrderNonDet>, \<Det> \} $} \\
\forallt{\kappa\ \beta} \<Det>\ \<List>\angles{\kappa\ \beta} \times \<Det Int>
& \strut\hspace{-1em} \rightarrow \<Det>\ \beta\phantom{\<Non>}
& \mbox{\quad in this case $\kappa = \<Det>$}
\end{eqnarray*}

@@

    \rulename{get}: 
    $
    \forallt{\tau} \kappa \in \{\<Det>, \<NonDet>\}\ \<OrderNonDet> \ C \angles{\tau} \ \times \kappa \ \<Int> \rightarrow 
    \<NonDet> \ \<Int>
    $

@@

    \rulename{set}: 
    $
    \forallt{\kappa, \beta} \ \<OrderNonDet> \ C \angles{\kappa \ \beta} \ \times \<Det> \ \<Int> \times \kappa \ \beta_n \rightarrow 
    \<NonDet> \ \beta
    $

@@

    \rulename{set invalid}: 
    $
    \forallt{\kappa, \beta} \ \<OrderNonDet> \ C \angles{\kappa \ \beta} \ \times \<NonDet> \ \<Int> \times \kappa \ \beta_n \rightarrow 
    \|INVALID|
    $

@@

    \rulename{hasnext element}:
    $
    \forallt{\tau} \<OrderNonDet>\ \<Iterator>\angles{\tau} \rightarrow
    \<Det> \ \<boolean>
    $
    \bigskip

>>>>

    \caption{Type well-formedness}

@@

The element type of a \<NonDet> collection must be \<NonDet>.
For other collections, the type qualifier on a collection's element must be a subtype or equal to
the collection's type qualifier (\rulename{collection} rules of
\cref{type-validity}). Note that the \rulename{ordernondet} rules also
permit \<Det> collections,
because $\<Det> \sqsubseteq \<OrderNonDet>$.
\Cref{fig:determinism-collections} restates the \rulename{collection} rules
of
\cref{type-validity}.

@@

\end{Verbatim}
\end{minipage}
\hfill
\begin{minipage}{.46\textwidth}
These types are invalid:
\begin{Verbatim}[commandchars=\\\{\}]
NonDet      \lptype{List}{OrderNonDet \lptype{Set}{...}}
NonDet      \lptype{List}{Det         Int}
OrderNonDet \lptype{List}{NonDet      Int}
Det         \lptype{List}{NonDet      Int}
Det         \lptype{List}{OrderNonDet \lptype{Set}{...}}
\end{Verbatim}
\end{minipage}
\smallskip

@@

\begin{enumerate}
\item
The individual elements retrieved from it have type \<NonDet>.  This
affects access, iteration, searching, etc.
(Rules \rulename{get}, \rulename{iteration}, and \rulename{next element} in \cref{fig-ordernondet-rules}.)
\item
Size-related operations return a deterministic result.  This also affects
queries of whether an iterator has more elements.
\item
If the collection is sorted, or its elements are placed in a collection
that does sorting, the result is deterministic.
\end{enumerate}

To state the first point more formally, the 
typical type for a list access operation, such as \<get>, is
$$\forallt{\tau} \<List>\angles{\tau} \times \<Int> \rightarrow \tau$$
but this type is incorrect for \ourTypeSystem.  Each of the following is a
correct (but overly restrictive) type for \<get>:
\begin{eqnarray*}
\forallt{\kappa\ \beta} \<NonDet>\ \<List>\angles{\kappa\ \beta} \times \<Det Int>
& \strut\hspace{-1em} \rightarrow \<NonDet>\ \beta
& \mbox{\quad in this case $\kappa = \<NonDet>$} \\
\forallt{\kappa\ \beta} \<OrderNonDet>\ \<List>\angles{\kappa\ \beta} \times \<Det Int>
& \strut\hspace{-1em} \rightarrow \<NonDet>\ \beta
& \mbox{\quad in this case $\kappa \in \{ \<OrderNonDet>, \<Det> \} $} \\
\forallt{\kappa\ \beta} \<Det>\ \<List>\angles{\kappa\ \beta} \times \<Det Int>
& \strut\hspace{-1em} \rightarrow \<Det>\ \beta\phantom{\<Non>}
& \mbox{\quad in this case $\kappa = \<Det>$}
\end{eqnarray*}

@@

    \rulename{get}: 
    $
    \forallt{\tau\ \kappa}
    % Not necessary: invalid instantiantions are not created
    % \in \{\<Det>, \<NonDet>\}\
    \<OrderNonDet> \ C \angles{\tau} \ \times \kappa \ \<Int> \rightarrow 
    \<NonDet> \ \<Int>
    $

@@

    \rulename{set}: 
    $
    \forallt{\kappa\ \beta} \ \<OrderNonDet> \ C \angles{\kappa \ \beta} \ \times \<Det> \ \<Int> \times \kappa \ \beta \rightarrow 
    \<NonDet> \ \beta
    $

@@

%% We don't need this.  Just leaving it out means the expression has no
%% typing and is therefore invalid.
%     \rulename{set invalid}: 
%     $
%     \forallt{\kappa, \beta} \ \<OrderNonDet> \ C \angles{\kappa \ \beta} \ \times \<NonDet> \ \<Int> \times \kappa \ \beta \rightarrow 
%     \|INVALID|
%     $

@@

    \rulename{hasnext element}:
    $
    \forallt{\tau} \<OrderNonDet>\ \<Iterator>\angles{\tau} \rightarrow
    \<Det> \ \<boolean>
    $
